Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism

### 2013-10-12 esop14/att-esop14.tex
Up through abs reps

\inferrule[e-op-kinding]{
	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvarCtx\\
	\kOpS{op}{\kappai} \in \Theta\\
	\tKindX{\taui}{\kappai}\\\\
	\exprOKX{e_1}\\
	\cdots\\
	\exprOKX{e_n}
}{
	\exprOKX{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}
}
\end{mathpar}
\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding rules for type-level terms are given in Figure \ref{tlkind}.}
\label{kindprog}
\end{figure}
\subsection{Active Typechecking and Translation}
The second phase of compilation involves typechecking and translating the program to produce an internal term, invoking the user-defined logic encoded in the introduced operator families when needed. The rules for active typechecking and translation are given in Fig. \ref{att}. The context $\fvalCtx$ tracks the operator definitions and representation schemas associated with families in scope. Because the logic inside operators must be invoked during this phase, we provide an evaluation semantics for type-level terms in Fig. \ref{tleval}.

The key rule is \textsc{att-exp}, which assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to an expression, $e$, as specified by the judgement $\ecompilesX{e}{\tau}{\gabs}$, then \emph{deabstracts} the translation to produce the final translation. Deabstraction is specified in Fig. \ref{deabs} by the judgement $\eraseX{\gabs}{\gamma}$. The purpose of the abstract translation is to ensure that the implementation details of a type cannot be observed by operators associated with a different type, ensuring that invariants maintained by that type cannot be violated by operators in a different type family. For example, the implementation of natural numbers as integers in Fig. \ref{example} maintains the invariant that the translation is non-negative. If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnum}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$.

This is prevented by a mechanism quite similar to the mechanism by which abstract types in ML-style modules operate \cite{atapl} -- by keeping the representation schema hidden outside the type family it is associated with. As a result, it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, and so $-1$ is not valid according to the rules we will describe below. The only way to produce a term of this type in an operator not associated with the type is to extract it from an input argument, where the invariants are guaranteed inductively to have been  maintained. We will make this more concrete in Sec. \ref{safety}.

Our representation schema abstraction mechanism is specified syntactically, so it relates to previous work on syntactic type abstraction \cite{syntypeabs}. While that work was focused on abstracting away the identity of a particular type outside of a ``host'', we focus on abstracting away the knowledge of how a primitive type family is implemented outside of the operators associated with it.

Variables translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$ (rule \textsc{att-var}). Lambda terms translate to lambda terms in the internal language.

\begin{figure}[t]
\small
$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtxX{\fvalDf}$
\begin{mathpar}
\inferrule[att-fam]{
	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
}{
	\pcompilesX{\pfam{\familyDf}{\rho}}
}

@@

\inferrule[att-lam]{
	\tEvalX{\tau_1}{\tau_1'}\\
	\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
}{
	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\sabs}{\gabs}}
}

\inferrule[att-op]{
	\fvalDf \in \fvalCtx\\
	\tops{op}{\kappai}{i}{a}{\tauop} \in \theta\\
	\tEvalX{\taui}{\taui'}\\\\
	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
	\cdots~~~~
	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\gamma_n}\\
	\text{$\begin{array}{c}
		\left[\text{$\begin{array}{r}
			\taui'/\tvar{i}\\
			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
			\end{array}$}\right]\tauop
		\Downarrow
		\tden{\titerm{\gamma}}{\ttype{Fam'}{\tauidx}}
		\end{array}$}\\
	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\sigma}\\
	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gamma}{\sigma}
}{
	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\gamma}
}
\end{mathpar}
\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}}
\label{att}
\end{figure}
\begin{figure}[t]
\small
%\vspace{-15pt}
$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
\begin{mathpar}
\small
\inferrule[var-kind]{
}{
  \tKind{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\tvar{t}}{\kappa}
}

>>>>

\inferrule[e-op-kinding]{
	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvarCtx\\
	\kOpS{op}{\kappai} \in \Theta\\
	\tKindX{\taui}{\kappai}\\\\
	\exprOKX{e_1}\\
	\cdots\\
	\exprOKX{e_n}
}{
	\exprOKX{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}
}
\end{mathpar}
\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding rules for type-level terms are given in Figure \ref{tlkind}.}
\label{kindprog}
\end{figure}
\subsection{Active Typechecking and Translation}
\begin{figure}[t]
\small
$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtxX{\fvalDf}$
\begin{mathpar}
\inferrule[att-fam]{
	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
}{
	\pcompilesX{\pfam{\familyDf}{\rho}}
}

@@

\inferrule[att-lam]{
	\tEvalX{\tau_1}{\tau_1'}\\
	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
}{
	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\gabs}}
}

\inferrule[att-op]{
	\fvalDf \in \fvalCtx\\
	\tops{op}{\kappai}{i}{a}{\tauop} \in \theta\\
	\tEvalX{\tauidx}{\tauidx'}\\\\
	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
	\cdots~~~~
	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\gamma_n}\\
	\text{$\begin{array}{c}
		\left[\text{$\begin{array}{r}
			\tauidx'/\tvar{i}\\
			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
			\end{array}$}\right]\tauop
		\Downarrow
		\tden{\titerm{\gabs}}{\tau}
		\end{array}$}\\
	\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}\\
	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gabs}{\sabs}
}{
	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\tauidx}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\gabs}
}
\end{mathpar}
\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}}
\label{att}
\end{figure}
The second phase of compilation involves typechecking and translating the program to produce an internal term, invoking the user-defined logic encoded in the introduced operator families when needed. The rules for active typechecking and translation are given in Fig. \ref{att}. The context $\fvalCtx$ tracks the operator definitions and representation schemas associated with families in scope. Because the logic inside operators must be invoked during this phase, we provide an evaluation semantics for type-level terms in Fig. \ref{tleval}.

The key rule is \textsc{att-exp}, which assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to an expression, $e$, as specified by the judgement $\ecompilesX{e}{\tau}{\gabs}$, then \emph{deabstracts} the translation to produce the final translation. Deabstraction is specified in Fig. \ref{deabs} by the judgement $\eraseX{\gabs}{\gamma}$. The purpose of the abstract translation is to ensure that the implementation details of a type cannot be observed by operators associated with a different type, which ensures that that invariants maintained by one type family are preserved even if other type families are subsequently introduced. For example, our implementation of natural numbers as integers in Fig. \ref{example} maintains the invariant that the translation is non-negative. If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnum}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$, breaking this invariant (and thus the guarantee that the recursor always terminates!)

To explain this mechanism, let us begin by going over how expressions are assigned types and abstract translations. Variables  (rule \textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$. Lambda terms (rule \textsc{att-lam}) translate to lambda terms in the internal language and are given the appropriate $\fvar{Arrow}$ type. The internal type assigned to the variable of the internal lambda term is $\trepof{\tau'}$, which is the abstract internal type associated with $\tau'$, the type of the variable in the lambda expression, $\tau$, after it has been fully evaluated to a type-level value. This will be converted to the actual internal type determined by $\tau'$'s representation schema during deabstraction.

This is prevented by a mechanism quite similar to the mechanism by which abstract types in ML-style modules operate \cite{atapl} -- by keeping the representation schema hidden outside the type family it is associated with. As a result, it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, and so $-1$ is not valid according to the rules we will describe below. The only way to produce a term of type $\ttype{Nat}{\tunit}$ in an operator not associated with the family $\fvar{Nat}$ is to extract it from an argument (e.g. by projecting it out of an $n$-tuple), so the invariants are guaranteed inductively to have been  maintained. We will make this non-interference guarantee more precise in Sec. \ref{safety}.

Our representation schema abstraction mechanism is purely syntactic, so it relates to previous work on syntactic type abstraction \cite{syntypeabs}. While that work was focused on abstracting away the identity of a particular type outside of the ``principal'' it is associated with, we focus on abstracting away the knowledge of how a primitive type family is implemented outside of its operators.

\begin{figure}[t]
\small
%\vspace{-15pt}
$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
\begin{mathpar}
\small
\inferrule[var-kind]{
}{
  \tKind{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\tvar{t}}{\kappa}
}
